To Batch or Not to Batch: Test-Ordering Variabilityin the ED

### 2024-01-25 manuscript/manuscript.tex
manuscript updates

\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the
Emergency Department and the Impact on Care Delivery}

\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the
Emergency Department and the Impact on Care Delivery}

@@

One important factor that can impact the speed and effectiveness of
diagnosis in the ED is the availability and performance of diagnostic
tests \citet{naseim2015}. A variety of diagnostic tests are used in the
ED, including laboratory tests, imaging studies, and specialized tests
such as electrocardiograms (ECGs) and point-of-care (POC) testing. These
tests can provide valuable information about a patient's condition and
help to guide treatment decisions.

@@

We start by evaluating the reduced-form effects of ED provider
batch-ordering tendency on downstream patient outcomes and turnaround
time. We find that practice variation as captured by physician
batch-ordering tendency has large and significant consequences. Being
treated by a provider in the top decile of the tendency distribution,
compared to being treated by someone in the bottom decile {[}INSERT
RESULTS{]}.

@@

Our analytical lens is focused on the Emergency Department at the Mayo
Clinic of Arizona, a distinguished tertiary care establishment. During
the timeframe of our study, the ED recorded an annual visitation of
approximately 45,000 patients. The department is singularly staffed by
board-eligible or board-certified emergency physicians, abstaining from
the services of nurse practitioners or physician assistants. A notable
observation was that residents in rotation oversaw a low fraction,
roughly 10\%, of the patient volume. Comprehensive patient data,
encompassing demographics, chief complaints, vital signs, emergency
severity, length of stay, and resource utilization metrics, were
meticulously logged during the study period.

@@

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe approximately \(48,000\) such visits until during
the study period. To improve power, we drop encounters with rare chief
complaints (\(<1000\) total encounters of this kind) and complaints
where a batch order occurs less than \(10\) percent of the time. Since
batch orders are rare for these cases, our physician batch tendency
instrument could suffer from a weak instrument problem if we included
them. Examples of complaints dropped include urinary complaints and
allergic reactions. Excluding these conditions does not introduce
selection bias unless physician test batching tendency is orthogonal to
physician diagnosing behavior. While this assumption may be violated if
we were to use a very detailed level of chief complaint information upon
which to base our exclusion criterion, it is plausibly satisfied when
using broad complaint
categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
In order to estimate a precise measure of physician-level batch
tendency, we further restrict our sample to the \(42,000\) encounters
involving physicians who treat over 500 ED cases per year.

@@

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice, where we
observe that approximately 97(?) percent of all tests ordered are
conducted. Below, we detail our primary outcomes: (a) ED length of stay
(LOS), (b) resource utilization, and (c) 72-hour return.

\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
efficiency and patient throughput in emergency care settings. It is
defined as the duration from a patient's arrival to the ED until their
departure, whether by discharge or admission to the hospital. Although
the literature provides a range for long ED-LOS, we operationalize long
ED-LOS as stays exceeding the 75th percentile in our dataset for this
study. This approach allows us to contextualize the definition of
extended stays according to the specific demographics and case mix of
the ED under study.

\emph{(b) Resource Utilization}-- Resource utilization in the ED
typically refers to the extent of medical services and interventions a
patient receives. In this study, we quantify resource utilization by the
total number of diagnostic tests ordered per patient during their ED
stay. This encompasses both initial and subsequent tests. The hypothesis
is that batch testing may lead to variations in the number of tests
ordered, potentially influencing the overall healthcare expenditure and
efficiency.

\emph{(c) 72-hour return}-- The rate of patients returning to the ED
within 72 hours of their initial visit is a proxy for the quality of
care received. A high 72-hour return rate might indicate inadequate
treatment or diagnostic oversight during the initial visit. We measure
this outcome as the proportion of patients with a subsequent ED visit
within 72 hours post-initial discharge.

@@

In the context of our study, batching in diagnostic test ordering is
defined as placing multiple test orders for a single patient within a
5-minute interval. To ensure the robustness of our analysis, sensitivity
tests were conducted on this cutoff point, affirming that our results
are consistent and reliable, irrespective of minor variations in the
batching definition.

Our study identifies and differentiates between two distinct strategies
of batching:

\begin{itemize}
\item
  Lab + Image Batch: This category of batch order includes instances
  where a patient's batched diagnostic tests comprise a combination of
  at least one imaging test (such as a contrasted CT scan,
  non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
  This mix is fascinating as it suggests a comprehensive diagnostic
  approach, possibly catering to complex medical cases.
\item
  Image + Image Batch: This category encompasses batch orders where a
  patient's diagnostic tests consist of two or more distinct imaging
  tests. The reliance solely on imaging in these batches may indicate
  specific diagnostic pathways or a focus on conditions that primarily
  require visual diagnostic methods.
\end{itemize}

It is important to note the differing implications of these batching
strategies. While lab tests often can be processed concurrently with
imaging tests, causing minimal additional waiting time, a batch of
multiple imaging tests requires the patient's presence for each test,
potentially leading to significant increases in waiting time. Thus, the
composition of a batch order can provide valuable insights into the
physician's decision-making process, particularly in balancing between
patient waiting times and diagnostic uncertainty. This balance can be
especially crucial in the context of varying complexity levels of
patient encounters.

Figure 1 illustrates the occurrence rates of each batching type across
the top ten most frequently reported chief complaints, providing a
visual representation of how batching practices vary with different
patient presentations.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Chief Complaint}
  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
\begin{tablenotes}
\tiny
\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
\end{tablenotes}  
\end{figure}

@@

In step two, the leniency measure for patient \(i\) seen by physician
\(j\) is computed as the average residual across all other patients seen
by the physician that year:

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure
\ref{fig:physician_batching} graphs the histogram of batch-ordering
frequency by physician for popular chief complaints, highlighting that
the variation in batching differs systematically.
\hyperref[tab:regression]{Table 3} presents the ``first stage'' in a
regression table: being assigned to a \(10\)pp higher batch-tendency
physician is associated with a \(8\)pp increase in the likelihood of
having tests batch-ordered in the ED. The F-statistic is 50 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the leniency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rare complaints.

@@

In Model 2 of Table \ref{tab:reducedform}, we control for testing
inclination and observe notable changes in the coefficients for batch
tendency. The previously significant positive association with ln(LOS)
diminishes, suggesting that the initial estimate may have been capturing
the combined effects of batching and testing inclination. Estimated
coefficients for batch tendency are scaled by the difference in batch
tendency going from the ninetieth to tenth percentile in physician batch
tendency for interpretability, and coefficients for testing inclination
are scaled by the difference in testing inclination going from the
ninetieth to tenth percentile in physician testing inclination.

Assignment to a physician in the top batching decile (relative to one in
the bottom decile) is associated with a 0.2 unit increase in the number
of diagnostic tests being ordered per patient encounter, without any
additional change in the likelihood of a 72-hour return. With an average
of 1,805 patient encounters per physician in our study period, this
translates to an additional 361 diagnostic tests with no perceived
benefit in terms of reducing returns or length of stay.

The effects of testing inclination are also striking. We find that a
movement from a physician in the top testing decile (relative to one in
the bottom decile) is associated with 0.6\% change in the length of stay
in the emergency department, a 0.96 unit increase in the number of
diagnostic tests being ordered per patient encounter, and -0.03 pp
change in the probability of a patient returning within 72 hours.

@@

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 10 percent of the time (recall that our
baseline sample only includes complaints with a \textgreater10 percent
batching rate). We estimate a reduced-form regression of each main
outcome on physician prescribing leniency for the subsample, following
equation (3). The results of this exercise are displayed in Table
\ref{tab:placebo_check}. They show that in contrast to results for our
main sample, the association between physician tendency to batch and a
given outcome is statistically indistinguishable from zero and much
smaller in magnitude for the samples of patients who visit the ED with
health conditions that are rarely batched.

@@

\hypertarget{investigating-mechanisms-excess-testing}{%
\subsection{Investigating Mechanisms: Excess
Testing}\label{investigating-mechanisms-excess-testing}}

In the previous section we find a particularly pronounced impact of
batching on resource utilization. We reported a significant coefficient
of 0.135 in the number of tests ordered due to batching, translating to
13.5 excess tests per 100 batch orders. Here, an ``excess test'' is
defined as one that would not have been ordered if the physician had
chosen to order tests sequentially. This suggests that the information
gain from sequential testing, where the need for additional tests is
informed by the results of earlier tests, could lead to more efficient
and targeted use of diagnostic resources.

To further understand this phenomenon, we probe into high-volume
complaint areas in the ED. By focusing on these areas, we can discern
which specific complaints are more likely to lead to excess testing
under a batch ordering system. This analysis is crucial as it sheds
light on the causal effect of batching in different clinical scenarios.

We consider the top 5 most common broad complaints in the ED, which
account for 82\% of all visits. Appendix table {[}\ldots{]} . These
include Cardiac/Chest-Related Issues, Gastrointestinal/Abdominal Issues,
Respiratory-Related Issues, Neurological/Syncope Issues, and
Musculoskeletal/Extremity Issues. We then examine the impact of batching
on the number of tests ordered for each of these complaints.

{[}\ldots{]}

@@

\begin{table}[!htbp]
\centering 
  \caption{Testing the Monotonicity Assumption} 
\label{tab:monotonicity}
\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
\hline \\[-1.8ex] 
Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample. Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

\begin{longtable}{|p{5cm}|p{12cm}|}
\caption{Condensed Complaints} \\
\hline
\textbf{Condensed Complaint Area} & \textbf{Complaints} \\
\hline
Gastrointestinal/Abdominal Issues & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia, Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
\hline
Cardiac/Chest-Related Issues & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia, Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
\hline
Musculoskeletal/Extremity Issues & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain, Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder, Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
\hline
Neurological/Syncope Issues & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor, Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertigo \\
\hline
Respiratory-Related Issues & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea, Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess\\
\hline
\end{longtable}

>>>>

\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variabilityin the ED}

\TITLE{To Batch or Not to Batch: Test-Ordering Variabilityin the ED}

@@

One important factor that can impact the speed and effectiveness of
diagnosis in the ED is the availability and performance of diagnostic
tests \citet{naseim2015}. A variety of diagnostic tests are used in the
ED, including laboratory tests, imaging studies, and specialized tests.
These tests can provide valuable information about a patient's condition
and help to guide treatment decisions.

@@

We start by evaluating the reduced-form effects of ED provider
batch-ordering tendency on downstream patient outcomes and turnaround
time. We find that practice variation as captured by physician
batch-ordering tendency has large and significant consequences. Being
treated by a provider in the top decile of the tendency distribution,
compared to being treated by someone in the bottom decile results in an
additional 2 diagnostic test per 100 patient encounters after
controlling for the physician's underlying propensity to test.

Physician care takes on multiple dimensions besides the decision to
batch order diagnostic tests. Employing a placebo exercise, we find
evidence that that excess testing occurs after having seen a
high-batching physician is due to batching from this physician, as
opposed to other differences across physician in care that correlate
with batching tendency: we find no effects of physician batch tendency
on key outcomes for a ``placebo sample'' of patients who visit the ED
for conditions that rarely result in batched orders.

With the caveat that other unobserved dimensions of physician care may
impact patient outcomes, we employ physician batch tendency as an
instrumental variable for having tests batch ordered in the ED to
quantify the effect of batching directly. We find that being treated by
a top decile batching physician, compared to a bottom decile batching
physician, results in an additional 13.5 tests per 100 patient
encounters. We find no evidence that batch-ordering reduces ED length of
stay or 72-hour return rates. These results suggest that the benefits of
batch-ordering are limited, and that the practice may be associated with
significant costs. Our findings are robust to analyses probing the key
assumptions of our batching definition.

@@

Our study was conducted in the Emergency Department (ED) of the Mayo
Clinic of Arizona, a tertiary care hospital without obstetrical
services, an inpatient pediatrics unit, or a trauma designation. During
the study period, the ED recorded approximately 43,000 visits per year,
managed across 26 treatment rooms and up to 9 hallway spaces. The
department is exclusively staffed by board-eligible or board-certified
emergency physicians (EPs), with rotating residents overseeing about
10\% of patient volume. Physicians operate in a unique workflow that
includes staggered 8.5-hour shifts and a rotational patient assignment
system, minimizing potential selection bias in patient encounters.

We conducted a retrospective review of comprehensive ED operational data
from 10/6/2018 through 12/31/2019, coinciding with the initiation of a
new electronic medical record. The dataset includes detailed patient
demographics, chief complaints, vital signs, emergency severity index
(ESI), length of stay (LOS), and resource utilization metrics. This
period was chosen to provide a robust data set while excluding the
influence of the coronavirus pandemic.

@@

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe approximately \(48,000\) such visits during the
study period. To improve power, we drop encounters with rare chief
complaints (\(<1000\) total encounters of this kind) and complaints
where a batch order occurs less than \(10\) percent of the time. Since
batch orders are rare for these cases, our physician batch tendency
instrument could suffer from a weak instrument problem if we included
them. Examples of complaints dropped include urinary complaints and
allergic reactions. Excluding these conditions does not introduce
selection bias unless physician test batching tendency is orthogonal to
physician diagnosing behavior. While this assumption may be violated if
we were to use a very detailed level of chief complaint information upon
which to base our exclusion criterion, it is plausibly satisfied when
using broad complaint
categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
In order to estimate a precise measure of physician-level batch
tendency, we further restrict our sample to the \(42,000\) encounters
involving physicians who treat over 500 ED cases per year.

@@

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice. Below, we
detail our primary outcomes: (a) ED length of stay (LOS), (b) resource
utilization, and (c) 72-hour return.

\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
efficiency and patient throughput in emergency care settings. It is
defined as the duration from a patient's arrival to the ED until their
departure, whether by discharge or admission to the hospital. The
hypothesis is that batch testing may lead to shorter ED-LOS, potentially
improving patient throughput and reducing crowding.

\emph{(b) Resource Utilization}-- Resource utilization in the ED
typically refers to the extent of medical services and interventions a
patient receives. In this study, we quantify resource utilization by the
total number of distinct diagnostic tests ordered per patient during
their ED stay. This encompasses both initial and subsequent tests. The
hypothesis is that batch testing may lead to variations in the number of
tests ordered, potentially influencing the overall healthcare
expenditure and efficiency.

\emph{(c) 72-hour return}-- The rate of patients returning to the ED
within 72 hours of their initial visit is a proxy for the quality of
care received. A high 72-hour return rate might indicate inadequate
treatment or diagnostic oversight during the initial visit. We measure
this outcome as the proportion of patients with a subsequent ED visit
within 72 hours post-initial discharge. The hypothesis is that batch
testing may lead to variations in the 72-hour return rate, potentially
influencing the quality of care received.

@@

We define ``batching'' in line with standard emergency medicine
practices. It involves a physician ordering a comprehensive set of
diagnostic tests at once, typically covering a broad range of potential
diagnoses. This contrasts with non-batching, where tests are ordered
more selectively based on the information available at the time, with
additional tests potentially ordered later as needed.

We operationalize batching as the ordering of multiple diagnostic tests
within a 5-minute window. Sensitivity analyses on this cutoff point
showed that our results are robust to this definition. This operational
definition aligns with the concept of batching as a single comprehensive
effort. For our analysis, all lab-based testing ordered in this
timeframe are categorized as one distinct type of testing (lab). In
contrast, each imaging test (e.g., X-ray, CT scan) are considered a
separate distinct test. Therefore, a batch in our study is defined as
consisting of two or more distinct diagnostic tests, which could be a
combination of lab and imaging tests or multiple imaging tests.

This approach reflects the realistic diagnostic strategies in an ED
setting and allows for a clear distinction between batching and
non-batching behaviors. The differentiation between lab tests as one
collective unit and imaging tests as individual units is based on their
operational impact in the ED, particularly regarding patient waiting
time and processing queues.

@@

In step two, the tendency measure for patient \(i\) seen by physician
\(j\) is computed as the average residual across all other patients seen
by the physician that year:

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure
\ref{fig:physician_batching} graphs the histogram of batch-ordering
frequency by physician for popular chief complaints, highlighting that
the variation in batching differs systematically.
\hyperref[tab:regression]{Table 3} presents the ``first stage'' in a
regression table: being assigned to a \(10\)pp higher batch-tendency
physician is associated with a \(8\)pp increase in the likelihood of
having tests batch-ordered in the ED. The F-statistic is 50 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the tendency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rare complaints.

@@

In Model 2 of Table \ref{tab:reducedform}, we control for testing
inclination and observe notable changes in the coefficients for batch
tendency. The previously significant positive association with ln(LOS)
diminishes, suggesting that the initial estimate may have been capturing
the combined effects of batching and testing inclination. Estimated
coefficients for batch tendency are scaled by the difference in batch
tendency going from the ninetieth to tenth percentile in physician batch
tendency for interpretability, and coefficients for testing inclination
are scaled by the difference in testing inclination going from the
ninetieth to tenth percentile in physician testing inclination.
Assignment to a physician in the top batching decile (relative to one in
the bottom decile) is associated with a 0.2 unit increase in the number
of diagnostic tests being ordered per patient encounter, without any
additional change in the likelihood of a 72-hour return.

The effects of testing inclination are also striking. We find that a
movement from a physician in the top testing decile (relative to one in
the bottom decile) is associated with 0.6\% change in the length of stay
in the emergency department, a 9.6 excess diagnostic tests being ordered
per 100 patient encounters, and -0.03 pp change in the probability of a
patient returning within 72 hours.

@@

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 10 percent of the time (recall that our
baseline sample only includes complaints with a \textgreater10 percent
batching rate). We estimate a reduced-form regression of each main
outcome on physician batch tendency for the subsample, following
equation (3). The results of this exercise are displayed in Table
\ref{tab:placebo_check}. They show that in contrast to results for our
main sample, the association between physician tendency to batch and a
given outcome is statistically indistinguishable from zero and much
smaller in magnitude for the samples of patients who visit the ED with
health conditions that are rarely batched.

@@

\hypertarget{robustness}{%
\subsection{Robustness}\label{robustness}}

The 2SLS results presented in the previous sections are robust to
several alternative specifications probing the key batching definition.
Our primary definition operationalizes batching as the ordering of
multiple diagnostic tests within a 5-minute window. This approach is
grounded in standard emergency medicine practices and offers a clear
distinction between batching and non-batching behaviors. To ensure the
robustness of our findings, we performed several sensitivity analyses.

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Variation in the Time Window for Batching: Firstly, we varied the time
  window for what constitutes a batch. The original 5-minute window was
  extended and contracted to understand its impact on the study's
  outcomes. Specifically, we considered scenarios where a batch is
  defined as two or more tests ordered within windows of 1 minute to 10
  minutes. This range allowed us to capture a broader spectrum of
  batching behaviors and test the sensitivity of our results to
  different batching definitions.
\item
  Refinement of Batching Definition: Secondly, we refined our batching
  definition based on the sequence and timing of test orders. In the
  initial analysis, if two tests were ordered upfront and another test
  ordered later, it was counted as a batch. We modified this definition
  to consider a set of tests as a batch only if all tests ordered during
  a patient encounter came from that initial batch order. This
  adjustment ensures that our batching definition more accurately
  reflects a comprehensive diagnostic effort at the outset of patient
  care, rather than incremental decision-making.
\end{enumerate}

The key finding from these robustness checks is the consistent impact of
batching on key outcomes across all variations. As seen in Table
\ref{tab:robustness_batching} and \ref{tab:robustness_batching2},
altering the time window for batching, whether narrowing it to as little
as 1 minute or expanding it to 10 minutes, did not qualitatively change
the results. Similarly, refining the definition of batching to consider
only those tests ordered in the initial batch also had negligible impact
on the study's main findings. These results lend further credibility to
our initial findings, demonstrating that our conclusions about the
impact of batching on patient outcomes in the ED are not sensitive to
the specific operational definition of batching.

@@

\begin{table}[!htbp]
\centering 
  \caption{Testing the Monotonicity Assumption} 
\label{tab:monotonicity}
\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
\hline \\[-1.8ex] 
Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\tiny
\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample. Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

\begin{table}[!htbp] \centering 
  \caption{2SLS Results: Robustness to Batching Definition (1-minute and 10-minute windows)} 
  \label{tab:robustness_batching}

\begin{tabular}{@{\extracolsep{10pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textbf{1-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
  & (0.203) & (0.018) & (0.009) \\ 
  & & & \\ 
 Batch & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
  & (0.496) & (0.054) & (0.023) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
Residual Std. Error & 0.492 & 0.836 & 0.184 \\ 
\hline 
\\[-1.8ex] & \multicolumn{3}{c}{\textbf{10-minute window}} \\
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Testing Inclination & 0.650$^{***}$ & 0.957$^{***}$ & $-$0.027$^{**}$ \\ 
  & (0.241) & (0.022) & (0.011) \\ 
  & & & \\ 
 Batch & $-$0.205 & 0.141$^{***}$ & 0.017 \\ 
  & (0.438) & (0.045) & (0.018) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.146 & 0.315 & 0.009 \\ 
Adjusted R$^{2}$ & 0.141 & 0.311 & 0.003 \\ 
Residual Std. Error & 0.512 & 0.814 & 0.184 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular}
\begin{tablenotes}
\tiny
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

\begin{table}[!htbp] \centering 
  \caption{2SLS Results: Robustness to Batching Definition (all tests at once)} 
  \label{tab:robustness_batching2}

\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Batch & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
  & (0.203) & (0.018) & (0.009) \\ 
  & & & \\ 
 `any.batch(fit)` & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
  & (0.496) & (0.054) & (0.023) \\ 
  & & & \\ 
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
Residual Std. Error (df = 41667) & 0.492 & 0.836 & 0.184 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
 & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
 & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
\end{tabular} 
\end{table}
